The \texttt{mc2d} package.

### 2010-03-02 pkg/mc2d/inst/doc/docmcEnglish.rnw
Some improvements on mc2d
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/riskassessment@85 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Solve some bugs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/riskassessment@89 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%\usepackage[letterpaper]{geometry}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}

@@

\begin{document}
\maketitle

This documentation is intended for readers with:
\begin{itemize}
\item A medium level of experience in R. Please refer to the Manual An
Introduction to R available with R distribution if needed;
\item Some knowledge about Monte-Carlo simulation (its basic principles
and its utility) and about Quantitative Risk Assessment (QRA). 
\end{itemize}

This documentation will not describe all arguments of the functions.
The definitive reference remains the documentation associated with
the package.

@@

\texttt{mc2d} means Two-Dimensional Monte-Carlo (\emph{Monte-Carlo
à Deux Dimensions}). This package :
\begin{itemize}
\item provides additional probability distributions;
\item provides tools to construct One-Dimensional and Two-Dimensional Monte-Carlo
Simulations;
\item provides tools to analyse One-Dimensional and Two-Dimensional Monte-Carlo
Simulations.
\end{itemize}

In a previous version, some tools to fit parametric distributions
to data were included. Because these functions are useful for other
purposes, they have been moved to a separate package called \texttt{fitdistrplus}.

\texttt{mc2d} was built for QRA in the Food Safety domain but it can
be used in other QRA domains.

@@

According to international recommendations, a QRA should reflect the
variability in the risk and calculate the uncertainty
associated with the risk estimate. The variability represents
temporal, geographical and/or individual heterogeneity of the risk
for a given population. The uncertainty is understood as stemming
from a lack of perfect knowledge about the QRA model structure and
associated parameters%
\footnote{In the engineering risk community, these concepts are refered as aleatoric
uncertainty for variability
and epistemic uncertainty for uncertainty.%
}.

In order to estimate the natural variability of the risk,
a Monte-Carlo simulation approach may be useful: the empirical distribution
of the risk within the population may be estimated from the mathematical
combination of distributions reflecting the variability of parameters
across the population.

A two-dimensional (or second-order) Monte-Carlo simulation was proposed
to estimate the uncertainty in the risk estimates stemming
from parameter uncertainty \cite{CULLEN-FREY-1999}. A two-dimensional
Monte-Carlo simulation is a Monte-Carlo simulation where the distributions
reflecting variability and the distributions
representing uncertainty are sampled
separately in the simulation, so that variability
and uncertainty in the output may be
estinated separately. It may be described as following (see Figure
\ref{fig:Mc2d}):

\begin{enumerate}
\item The parameters of the model should be divided into three categories:
the parameters whose distributions reflect variability only,
hereinafter denoted as variable parameters, the parameters
whose distributions reflect uncertainty only, denoted as uncertain
parameters and the parameters whose distributions reflect both
uncertainty and variability. For this latter category, a hierarchical
structure, using hyper-parameters, should be specified: if
a parameter is both uncertain and variable, one should be able to
specify an empirical or parametric distribution representing variability.
This distribution is conditional upon other parameters for which there
is some associated uncertainty. As an example, one should be able
to specify a relationship such as
\[ X\left|a,b\right.\sim N\left(a,b\right) \]
where the specified normal distribution represents variability in
$x$ conditional upon parameters $a$ and $b$. Hyperdistributions,
such as
\[ a\sim Unif\left(l_{a},u_{a}\right) \]
and
\[ b\sim Unif\left(l_{b},u_{b}\right) \]
represent the uncertainty in the parameters $a$ and $b$;
\item A set of uncertain parameters are randomly sampled from their respective
distributions; 
\item The QRA is performed using a classical (one-dimensional) Monte-Carlo
simulation of size $N_{v}$, treating the uncertain parameters as
fixed. This QRA takes into account the variability in all variable
parameters, and leads to an empirical density function reflecting
the variability of exposure/risk across the population, conditional
upon the uncertain parameters. Various statistics (\emph{e.g.} the
mean, the standard deviation, percentiles) of the resulting empirical
density function are evaluated and stored; 
\item Steps 2) and 3) are performed a large number ($N_{u}$) of times ,
leading to $N_{u}$ sets of statistics; 
\item As output, the $50^{th}$ percentile (median) of each statistic
is used as a point estimate of this statistic; the $2.5^{th}$
and $97.5^{th}$ percentiles of each statistic are used to
establish a 95\% credible interval (CI95) of this statistic. The median
of the $N_{u}$ estimated values for each of the 101 estimated percentiles
allows us to display a variability cumulative distribution
via a graph. This curve is surrounded by the 2.5th and 97.5th percentiles
obtained from the $N_{u}$ estimates of each of the 101 percentiles.
\end{enumerate}
%
\begin{figure}
\noindent \begin{centering}
\includegraphics[angle=270, width=1\textwidth]{mc2dsaumon}
\par\end{centering}

@@

\texttt{mc2d} is a set of R functions
that will help to develop such two-dimensional Monte-Carlo simulations.
The main difference from the procedure described above is that \texttt{mc2d}
uses arrays of (at least) two dimensions to derive the results: the
first dimension will reflect variability, the second will reflect
uncertainty. This document will not develop the method further, but
will illustrate the practical application of \texttt{mc2d}, using
a fictitious example.

@@

\textbf{Quantitative Risk Assessment: \emph{Escherichia coli}
O157:H7 infection linked to the consumption of frozen ground
beef in <3 year old children.}
\begin{itemize}
\item We assume that, in a given batch of ground beef, \emph{E. coli} O157:H7
are randomly distributed with a mean concentration of $c=10$ bacteria
(cfu) per gram of product;
\item We assume that no bacterial growth occurs in storage, since the product
is kept frozen until it is cooked, just before consumption;
\item 2.7\% of consumers cook their beef rare, 37.3\% medium
and 60.0\% well done; 
\item The following bacterial inactivation $i$ is associated with these
cooking practices:

\begin{itemize}
\item No inactivation for rare cooking;
\item 1/5 surviving bacteria for a medium cooking;
\item 1/50 surviving bacteria for a well done cooking.
\end{itemize}
\item The variability in steak serving sizes $s$ for <3 year children was
estimated in a consumption survey. The best fit to the data
was a gamma distribution with parameters: $shape=3.93,\: rate=0.0806$.
\item The dose-response relationship, describing the probability of illness,
$P,$ according to the dose is a one-hit model. The probability of
illness per hit $r$ is assumed to be constant with $r=0.001$.
\end{itemize}
The question is: \emph{What is the distribution of the risk of
illness in the population that consumed the contaminated lot?}

@@

As a first step, we assume that no uncertainty exists in our model.
All distributions represent variability only. The model may be written
as:
\begin{eqnarray*}
c & = & 10.\\
i & \sim & emp\left(\left\{ 1,1/5,1/50\right\} ,\left\{ 0.027,0.373,0.600\right\} \right)\\
s & \sim & gamma\left(3.93,0.0806\right)\\
n & \sim & Poisson\left(c\times i\times s\right)\\
P & = & 1-\left(1-0.001\right)^{n}
\end{eqnarray*}

@@

%true
<<sh1, eval=true, echo=true>>=
library(mc2d)
ndvar(1000)
conc <- 10
cook <- mcstoc(rempiricalD, values=c(1,1/5,1/50), prob=c(0.027,0.373,0.600))
serving <- mcstoc(rgamma,shape=3.93,rate=0.0806)
expo <- conc * cook * serving
dose <- mcstoc(rpois,lambda=expo)
r <- 0.001
risk <- 1-(1-r)^dose
EC1 <- mc(cook,serving,expo,dose,risk)
print(EC1)
summary(EC1)
@

@@

Note that the distributions of $r$ and $c$ represent uncertainty,
while the distributions of $i$ and $s$ represent variability. $n$,
which is a function of $c$, $i$ and $s,$ will be both variable
\emph{and} uncertain.

@@

Note that the syntax is similar to the earlier model. However, a \texttt{type}
argument is provided for each distribution, indicating whether the
parameter distribution represents uncertainty (\texttt{type}=\texttt{U}),
variability (\texttt{type}=V), or both
(\texttt{type}=\texttt{VU}).

@@

A typical session of R using \texttt{mc2d} is as follows:
\begin{itemize}
\item From data, expert knowledge, \emph{etc.} an empirical or parametric
distribution is chosen for each parent parameter. For developing
an empirical distribution from data , the \texttt{fitdistrplus}
package is recommended;
\item For each parameter, an \texttt{mcnode} object is constructed (key
functions: \texttt{mcdata, mcstoc}); 
\item Various \texttt{mcnode} objects are grouped into an \texttt{mc} object
(key function: \texttt{mc}).
\item The \texttt{mc} object is studied through summaries, graphs, and sensitivity
analysis (key functions: \texttt{summary.mc}, \texttt{plot.mc}, \texttt{tornado},
\texttt{tornadounc}).
\end{itemize}

@@

An \texttt{mcnode} object is the basic element of an \texttt{mc} object.
It is an array of dimension \texttt{nsv} * \texttt{nsu} * \texttt{nvariates}
where $nsv$ is the dimension of variability, $nsu$ is the dimension
of uncertainty and $nvariates$ is the number of variates of the \texttt{mcnode}%
\footnote{In this section, we will only consider \texttt{mcnode}s with $nvariates=1$.%
}. Four types of \texttt{mcnode} exist: 
\begin{itemize}
\item \texttt{V} \texttt{mcnode}, for \emph{Variability},
is an array of dimension $\left(nsv\times1\times nvariates\right)$.
The distribution represents variability in the parameter; 
\item \texttt{U} \texttt{mcnode}, for \emph{Uncertainty},
is an array of dimension $\left(1\times nsu\times nvariates\right)$.
The distribution represents uncertainty in the parameter. 
\item \texttt{VU} \texttt{mcnode}, for \emph{Variability
and Uncertainty}, is an array of dimension $\left(nsv\times nsu\times nvariates\right)$.
The distribution represents both variability (in the first dimension)
and uncertainty (in the second dimension) in the parameter. 
\item Additionally, a \texttt{0} \texttt{mcnode}
is also defined . \texttt{0} stands
for Neither Variability or Uncertainty.
Such nodes are arrays of dimension $\left(1\times1\times nvariates\right)$.
No uncertainty or variability is considered for these nodes. A \texttt{0}
\texttt{mcnode} is not necessary in the univariate context (use a
scalar instead) but is useful in constructing multivariate nodes (See
section \ref{sec:Multivar}). 
\end{itemize}
%
\begin{figure}
\noindent \begin{centering}
\includegraphics[angle=270, width=.5\textwidth]{Illustration}
\par\end{centering}

@@

There are 5 ways to construct an \texttt{mcnode} object:
\begin{enumerate}
\item The \texttt{mcstoc} function constructs an \texttt{mcnode} from random
number generating functions;
\item The \texttt{mcdata} function constructs an \texttt{mcnode} from data
sets; 
\item \texttt{An mcnode} can be constructed directly from operations on
\texttt{mcnode} objects;
\item \texttt{mcprobtree} is a special function that constructs an \texttt{mcnode}
from other \texttt{mcnodes} using a probability tree;
\item Some functions, such as \texttt{==} or \texttt{>}
, \texttt{is.na}, \texttt{is.finite} generate a new \texttt{mcnode}
when applied to an existing \texttt{mcnode}.
\end{enumerate}

@@

The \texttt{mcstoc} function is written as%
\footnote{as is standard in R, most arguments have logical default values and
will be infrequently modified.%
}:

%\begin{singlespace}
\noindent \begin{flushleft}
\texttt{mcstoc(func=runif, type=c(V,
U, VU,
0), ..., nsv=ndvar(), nsu=ndunc(), nvariates=1,
outm=each, nsample=n,
seed=NULL, rtrunc=FALSE, linf=-Inf, lsup=Inf, lhs=FALSE)}
\par\end{flushleft}
%\end{singlespace}
\begin{itemize}
\item \texttt{func} is a function providing random data or its name as a
character. The table \ref{tab:Distributions} provides available distributions
from the \texttt{stats} and the \texttt{mc2d} libraries that can be
used in \texttt{mcstoc}; 
\item \texttt{type} is the type of requested \texttt{mcnode}. By default,
\texttt{mcstoc} constructs a \texttt{V} \texttt{mcnode};
\item \ldots{} are the arguments to be passed to the function \texttt{func},
with the exception of the argument providing the size of the sample.
This latter is calculated by the function according to \texttt{func},
\texttt{type}, \texttt{nsv,} \texttt{nsu} and \texttt{nvariates}.
If the name of the argument specifying the size of the sample is not
\texttt{n} (\emph{e.g.} functions \texttt{rhyper} and \texttt{rwilcox},
see table \ref{tab:Distributions}), the name of this parameter should
be provided in the \texttt{nsample} argument. \emph{Note that all
of the following arguments should be named};
\item \texttt{nsv} and \texttt{nsu} are the number of samples needed in
the variability and uncertainty dimension, respectively. By default,
these values are the ones provided by \texttt{ndvar()} and \texttt{ndunc()},
respectively;
\item \texttt{nvariates} is the desired number of variates in the \texttt{mcnode};
\item \texttt{outm} is the default output for multivariate nodes;
\item \texttt{seed} optionally specifies a seed for the random number generator;
\item \texttt{rtrunc} allows truncation of a distribution between \texttt{linf}
and \texttt{lsup}. This function is not valid for every distribution
(see table \ref{tab:Distributions}). See the \texttt{rtrunc} function
help for further details;
\item \texttt{lhs} allows Latin hypercube sampling of the node . This function
is not valid for every distribution (see table \ref{tab:Distributions}).
See the \texttt{lhs} function help for further details. 
\end{itemize}
%
\begin{table}
\caption{\label{tab:Distributions}Available distributions}

@@

A normal distribution with parameters $mean=2$, $sd=3$, truncated
on the interval {[}1.5, 2{]}, with samples generated via Latin hypercube
sampling could be written%
\footnote{Note that the mean and the standard deviation of the untruncatednormal
distribution are not preserved in the truncated distribution.%
}:

@@

For convenience in using \texttt{mcstoc}, the following additional
distributions have been implemented: the Bernoulli distribution (\texttt{rbern}),
the empirical discrete distribution (\texttt{rempiricalD}), the PERT
distribution (\texttt{rpert})\cite{VOSE-2000}, the triangular distribution
(\texttt{rtriang}), the Dirichlet distribution (\texttt{rdirichlet})
and the multivariate normal distribution (\texttt{rmultinormal}).
The multinomial distribution has been adapted (vectorized): \texttt{rmultinomial}
(library \texttt{mc2d}) should be used in place of \texttt{rmultinom}
(library \texttt{stats}). The empirical discrete (\emph{e.g.} for
bootstrap), the Dirichlet, the multinomial and the multivariate normal
may be used with uncertain and/or variable parameters by specifying
multivariate nodes. See section \ref{sec:Multivar}.

@@

Another way to construct an \texttt{mcnode} object is \emph{via} the
\texttt{mcdata} function, when data are available.

@@

\texttt{mcnode}s can be automatically constructed using operations
on other \texttt{mcnode}s. Rules are used to transfer uncertainty
and variability coherently within the model. Logically, the rules
are as follows (illustrated here with a +)%
\footnote{These rules are not the standard R rules for recycling.%
}:
\begin{itemize}
\item \texttt{0} + \texttt{0}
= \texttt{0};
\item \texttt{0} + \texttt{V}
= \texttt{V}
\item \texttt{0} + \texttt{U}
= \texttt{U};
\item \texttt{0} + \texttt{VU}
= \texttt{VU};
\item \texttt{V} + \texttt{V}
= \texttt{V};
\item \texttt{V} + \texttt{U}
= \texttt{VU}: the \texttt{U}
mcnode is recycled by row, the \texttt{V}
mcnode is recycled in the standard manner by column;
\item \texttt{V} + \texttt{VU}
= \texttt{VU}: the \texttt{V}
mcnode is recycled in the standard manner by column;
\item \texttt{U} + \texttt{U}
= \texttt{U};
\item \texttt{U} + \texttt{VU}
= \texttt{VU}: the \texttt{U}
mcnode is recycled by row;
\item \texttt{VU} + \texttt{VU}
= \texttt{VU}
\end{itemize}
Thus, in our example:

@@

\texttt{expo} is a function of a \texttt{U} and two \texttt{V}
\texttt{mcnode}s: it is a \texttt{VU} \texttt{mcnode} with
variability in the row dimension and uncertainty in the column dimension
. \texttt{risk} is a function of a \texttt{U} and a \texttt{VU}
\texttt{mcnode}: it is therefore a \texttt{VU} \texttt{mcnode}.

@@

Special functions \texttt{is.na(x)}, \texttt{is.nan(x)}, \texttt{is.finite(x)},
\texttt{is.infinite(x)} are implemented to test if any values are
\texttt{NA} (missing data), \texttt{NaN} (\emph{Not A Number}),
or finite .

@@

Note that the resulting correlation (around 0.4) is obviously an approximation
to the desired value of 0.5, because a discrete distribution (\texttt{cook}:
3 categories) is correlated with a continuous distribution (\texttt{serving}).

@@

One can re-run the model with various dimensions or random seeds in
one line:

@@

Please refer any comments or bugs to rpouillot@yahoo.fr.

>>>>

%\usepackage[letterpaper]{geometry}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[scale=0.8, centering]{geometry}

@@

\begin{document}

\newpage
\maketitle
\newpage

This documentation is intended for readers with:
\begin{itemize}
\item A medium level of experience in R. Please refer to the Manual \emph{An
Introduction to R} available with R distribution if needed;
\item Some knowledge about Monte-Carlo simulation (its basic principles
and its use) and about Quantitative Risk Assessment (QRA). 
\end{itemize}

This documentation will not describe all arguments of the functions.
The definitive reference remains the documentation associated with
the package. It is recommended to try the examples while reading the 
explanations.

@@

\texttt{mc2d} means Two-Dimensional Monte-Carlo (\emph{Monte-Carlo
à 2 Dimensions}). This package provides:
\begin{itemize}
\item additional probability distributions;
\item tools to construct One-Dimensional and Two-Dimensional Monte-Carlo
Simulations;
\item tools to analyse One-Dimensional and Two-Dimensional Monte-Carlo
Simulations.
\end{itemize}

In a previous version, some tools to fit parametric distributions
to data were included. Because these functions can be useful for other
purposes, they have been moved to a separate package called \texttt{fitdistrplus}.

\texttt{mc2d} was built for QRA in the Food Safety domain but it can
be used in other frameworks.

@@

According to international recommendations, a QRA should reflect the
variability in the risk and take into account the uncertainty
associated with the risk estimate. The variability represents
for instance temporal, geographical and/or individual heterogeneity of the risk
for a given population. The uncertainty is understood as stemming
from a lack of perfect knowledge about the QRA model structure and
associated parameters%
\footnote{In the engineering risk community, these concepts are refered as aleatoric
uncertainty for variability
and epistemic uncertainty for uncertainty.%
}.

In order to reflect the natural variability of a modelled risk,
a Monte-Carlo simulation approach may be useful: the empirical distribution
of the risk within the population may be obtained from the mathematical
combination of distributions reflecting the variability of parameters
across the population.

A two-dimensional (or second-order) Monte-Carlo simulation was proposed
to superimpose the uncertainty in the risk estimates stemming
from parameter uncertainty \cite{CULLEN-FREY-1999}. A two-dimensional
Monte-Carlo simulation is a Monte-Carlo simulation where the distributions
reflecting variability and the distributions
representing uncertainty are sampled
separately in the simulation, so that variability
and uncertainty in the output may be
assessed separately. It may be described as following (see Figure
\ref{fig:Mc2d}):

\begin{enumerate}
\item The parameters of the model should be divided into four categories:
the fix parameters, the parameters whose distributions reflect variability only,
hereinafter denoted as \emph{variable parameters}, the parameters
whose distributions reflect uncertainty only, denoted as \emph{uncertain
parameters} and the parameters whose distributions reflect both
uncertainty and variability, denoted as \emph{variable and uncertain parameters}.
For this latter category, a hierarchical
structure, using hyper-parameters, has to be specified: if
a parameter is both uncertain and variable, one should be able to
specify an empirical or parametric distribution representing variability.
This distribution is conditional upon other parameters to which is
associated the uncertainty. As an example, one should be able
to specify a relationship such as
\[ R\left|a,b\right.\sim N\left(a,b\right) \]
where the specified normal distribution represents variability in
$r$ conditional upon parameters $a$ and $b$. Hyperdistributions,
such as
\[ a\sim Unif\left(l_{a},u_{a}\right) \]
and
\[ b\sim Unif\left(l_{b},u_{b}\right) \]
represent the uncertainty in the parameters $a$ and $b$ with uniform distribution;
\item A set of uncertain parameters are randomly sampled from their respective
distributions; 
\item The model is evaluated using a classical (one-dimensional) Monte-Carlo
simulation of size $N_{v}$, treating the uncertain parameters as
fixed. This QRA takes into account the variability in all variable
parameters, and leads to an empirical density function reflecting
the variability of exposure/risk across the population, conditional
upon the uncertain parameters. Various statistics (\emph{e.g.} the
mean, the standard deviation, percentiles) of the resulting empirical
density function are evaluated and stored; 
\item Steps 2) and 3) are performed a large number ($N_{u}$) of times ,
leading to $N_{u}$ sets of statistics. The uncertain parameters are generated
with respect to their uncertain distributions; 
\item As output, the $50^{th}$ percentile (median) of each statistic
is used as a point estimate of this statistic; the $2.5^{th}$
and $97.5^{th}$ percentiles of each statistic are used to
establish a 95\% credible interval (CI95) of this statistic. The median
of the $N_{u}$ estimated values for each of the 101 estimated percentiles
allows us to display a variability cumulative distribution
via a graph. This curve is surrounded by the 2.5th and 97.5th percentiles
obtained from the $N_{u}$ estimates of each of the 101 percentiles.
\end{enumerate}

\begin{figure}
\noindent \begin{centering}
\includegraphics[angle=270, width=1\textwidth]{mc2dsaumon}
\par\end{centering}

@@

\texttt{mc2d} is a set of R functions
that will help to implement such two-dimensional Monte-Carlo simulations.
The main point to understand is that \texttt{mc2d}
uses arrays of (at least) two dimensions to derive the results: the
first dimension will reflect variability, the second will reflect
uncertainty. This document will not develop the method further, but
will illustrate the practical application of \texttt{mc2d}, using
a fictitious example.

@@

\textbf{Quantitative Risk Assessment: \emph{Escherichia coli}
O157:H7 infection linked to the consumption of frozen ground
beef in <3 year old children.}

The model is built assuming that:
\begin{itemize}
\item in a given batch of ground beef, \emph{E. coli} O157:H7
are randomly distributed with a mean concentration of $c=10$ bacteria
(cfu) per gram of product;
\item no bacterial growth occurs in storage, since the product
is kept frozen until it is cooked, just before consumption;
\item 2.7\% of consumers cook their beef rare, 37.3\% medium
and 60.0\% well done; 
\item The following bacterial inactivation $i$ is associated with these
cooking practices: 
\begin{itemize}
\item No inactivation for rare cooking;
\item 1/5 surviving bacteria for a medium cooking;
\item 1/50 surviving bacteria for a well done cooking.
\end{itemize}
\item The variability in steak serving sizes $s$ for <3 year children can 
be described with a gamma distribution with parameters: $shape=3.93,\: rate=0.0806$.
\item The dose-response relationship, describing the probability of illness,
$P,$ according to the dose is a one-hit model. The probability of
illness per hit $r$ is assumed to be constant with $r=0.001$.
\end{itemize}
The question is: \emph{What is the risk of
illness in the population that consumed the contaminated lot?}

@@

As a first step, we assume that no uncertainty exists in our model.
All distributions represent variability only. The model may be written
as:
\begin{eqnarray*}
c & = & 10.\\
i & \sim & emp\left(\left\{ 1,1/5,1/50\right\} ,\left\{ 0.027,0.373,0.600\right\} \right)\\
s & \sim & gamma\left(3.93,0.0806\right)\\
n & \sim & Poisson\left(c\times i\times s\right)\\
r & = & 0.001\\
P & = & 1-\left(1-r\right)^{n}
\end{eqnarray*}

@@

%true
<<sh1, eval=true, echo=true>>=
library(mc2d)
ndvar(1001)
conc <- 10
cook <- mcstoc(rempiricalD, values=c(1,1/5,1/50), prob=c(0.027,0.373,0.600))
serving <- mcstoc(rgamma,shape=3.93,rate=0.0806)
expo <- conc * cook * serving
dose <- mcstoc(rpois,lambda=expo)
r <- 0.001
risk <- 1-(1-r)^dose
EC1 <- mc(cook,serving,expo,dose,risk)
print(EC1)
summary(EC1)
@

The \texttt{print} output provides for each \texttt{node}: 
the \texttt{mode} of the variable (numeric or logical), 
\texttt{nsv}, the size of the variability dimension (here, 1000), 
\texttt{nsu}, the size of the uncertainty dimension (here, 1 since no uncertainty is considered),
\texttt{nva}, the number of variate (here, all the variables are univariate),
some statistics such as the \texttt{min}imum, the \texttt{mean}, 
the \texttt{median} and the \texttt{max}imum,
evaluated after gathering the two dimensions,
the number of missing data \texttt{Nas}, 
the \texttt{type} of the variable (\texttt{0} for fix,
\texttt{V} for variable, \texttt{U} for uncertain or \texttt{VU} for variable
and uncertain (here, all nodes are variable parameters only)
and, finally, the level of output \texttt{outm}, for multivariate parameters.
The \texttt{summary} output proovides additionnal statistics.

@@

Note that the distributions of $r$ and $c$ represent uncertainty,
while the distributions of $i$ and $s$ represent variability. $n$,
which is a function of $c$, $i$ and $s,$ will become both variable
\emph{and} uncertain.

@@

Note that the syntax is similar to the earlier model. However, a \texttt{type}
argument is provided for each distribution, indicating whether the
parameter distribution represents variability (\texttt{type=V}, by default),
uncertainty (\texttt{type=U}),
 or both (\texttt{type=VU}).

@@

A typical session of R using \texttt{mc2d} is as follows:
\begin{itemize}
\item From data, expert knowledge, \emph{etc.} an empirical or parametric
distribution is chosen for each parent parameter. For developing
a parametric distribution from data, the \texttt{fitdistrplus}
package is a convenient way;
\item For each parameter, an \texttt{mcnode} object is constructed (key
functions: \texttt{mcdata, mcstoc}); 
\item Various \texttt{mcnode} objects are grouped into an \texttt{mc} object
(key function: \texttt{mc}).
\item The \texttt{mc} object is studied through summaries, graphs, and sensitivity
analysis (key functions: \texttt{summary.mc}, \texttt{plot.mc}, \texttt{tornado},
\texttt{tornadounc}).
\end{itemize}

@@

An \texttt{mcnode} object is the basic element of an \texttt{mc} object.
An \texttt{mcnode}  is associated to one variable while an \texttt{mc} is 
a set of associated variables.

An \texttt{mcnode} is an array of dimension \texttt{nsv} * \texttt{nsu} * \texttt{nvariates}
where $nsv$ is the dimension of variability, $nsu$ is the dimension
of uncertainty and $nvariates$ is the number of variates of the \texttt{mcnode}%
\footnote{In this section, we will only consider univariate \texttt{mcnode}s,
that is \texttt{mcnode}s with $nvariates=1$.%
}. Four types of \texttt{mcnode} exist: 
\begin{itemize}
\item \texttt{V} \texttt{mcnode}, for \emph{Variability},
is an array of dimension $\left(nsv\times1\times nvariates\right)$.
The distribution represents variability in the parameter; 
\item \texttt{U} \texttt{mcnode}, for \emph{Uncertainty},
is an array of dimension $\left(1\times nsu\times nvariates\right)$.
The distribution represents uncertainty in the parameter. 
\item \texttt{VU} \texttt{mcnode}, for \emph{Variability
and Uncertainty}, is an array of dimension $\left(nsv\times nsu\times nvariates\right)$.
The distribution represents both variability (in the first dimension)
and uncertainty (in the second dimension) in the parameter. 
\item Additionally, a \texttt{0} \texttt{mcnode}
is also defined . \texttt{0} stands
for Neither Variability or Uncertainty.
Such nodes are arrays of dimension $\left(1\times1\times nvariates\right)$.
No uncertainty or variability is considered for these nodes. A \texttt{0}
\texttt{mcnode} is not necessary in the univariate context (use a
scalar instead) but is useful in constructing multivariate nodes (See
section \ref{sec:Multivar}). 
\end{itemize}
%
\begin{figure}
\noindent \begin{centering}
\includegraphics[angle=270, width=.7\textwidth]{Illustration}
\par\end{centering}

@@

There are several ways to construct an \texttt{mcnode} object:
\begin{enumerate}
\item The \texttt{mcstoc} function constructs an \texttt{mcnode} from random
number generating functions;
\item The \texttt{mcdata} function constructs an \texttt{mcnode} from data
sets; 
\item \texttt{An mcnode} can be constructed directly from operations on
\texttt{mcnode} objects;
\item \texttt{mcprobtree} is a special function that constructs an \texttt{mcnode}
from other \texttt{mcnodes} using a  mixture of distribution, called ''probability tree''
in QRA;
\item Some functions, such as \texttt{==} or \texttt{>}
, \texttt{is.na}, \texttt{is.finite} generate a new \texttt{mcnode}
when applied to an existing \texttt{mcnode}.
\end{enumerate}

@@

The \texttt{mcstoc} function \footnote{from \texttt{stoc}hastic)} is written as
\footnote{as is standard in R, most arguments have default values and
will be infrequently modified.}:

%\begin{singlespace}
\noindent \begin{flushleft}
\texttt{mcstoc(func=runif, type=c(V,
U, VU,
0), ..., nsv=ndvar(), nsu=ndunc(), nvariates=1,
outm=each, nsample='n',
seed=NULL, rtrunc=FALSE, linf=-Inf, lsup=Inf, lhs=FALSE)}
\par\end{flushleft}
%\end{singlespace}
\begin{itemize}
\item \texttt{func} is a function providing random data or its name as a
character. The table \ref{tab:Distributions} provides available distributions
from the \texttt{stats} and the \texttt{mc2d} libraries that can be
used in \texttt{mcstoc}; 
\item \texttt{type} is the type of requested \texttt{mcnode}. By default,
\texttt{mcstoc} constructs a \texttt{V} \texttt{mcnode};
\item \ldots{} are the arguments to be passed to the function \texttt{func},
with the exception of the argument providing the size of the sample.
This latter is calculated by the function according to \texttt{func},
\texttt{type}, \texttt{nsv}, \texttt{nsu} and \texttt{nvariates}.
\emph{Note that all of the following arguments should be named};
\item \texttt{nsv} and \texttt{nsu} are the number of samples needed in
the variability and uncertainty dimensions, respectively. By default,
these values are the ones provided by \texttt{ndvar()} and \texttt{ndunc()},
respectively;
\item \texttt{nvariates} is the desired number of variates in the \texttt{mcnode}. 
see section \ref{sec:Multivar};
\item \texttt{outm} is the default output for multivariate nodes. see section \ref{sec:Multivar};
\item \texttt{nsample} is the name of the argument of \texttt{func} specifying 
the size of the sample. It is usually \texttt{n}, with the notable exception of
the \texttt{rhyper} and \texttt{rwilcox} where \texttt{nsample} should be 
changed in \texttt{nn}.
\item \texttt{seed} optionally specifies a seed for the random number generator.
If \texttt{NULL}, the seed is unchanged;
\item \texttt{rtrunc} allows truncation of a distribution between \texttt{linf}
and \texttt{lsup}. This option is not valid for every distribution
(see table \ref{tab:Distributions}). See the \texttt{rtrunc} function
help for further details;
\item \texttt{lhs} allows Latin hypercube sampling of the node . This function
is not valid for every distribution (see table \ref{tab:Distributions}).
See the \texttt{lhs} function help for further details. 
\end{itemize}
%
\begin{table}
\caption{\label{tab:Distributions}Available distributions}

@@

A normal distribution with parameters $mean=2$, $sd=3$, truncated
on the interval {[}1.5, 2{]}, with samples generated via Latin hypercube
sampling could be written%
\footnote{Note that the mean and the standard deviation of the untruncated normal
distribution are not preserved in the truncated distribution.%
}:

@@

For convenience in using \texttt{mcstoc}, the following additional
distributions have been implemented: the Bernoulli distribution (\texttt{rbern}),
the empirical discrete distribution (\texttt{rempiricalD}), the PERT
distribution (\texttt{rpert})\cite{VOSE-2000}, the triangular distribution
(\texttt{rtriang}), the Dirichlet distribution (\texttt{rdirichlet})
and the multivariate normal distribution (\texttt{rmultinormal}).
The multinomial distribution has been adapted (vectorized) and \texttt{rmultinomial}
(library \texttt{mc2d}) should be used in place of \texttt{rmultinom}
(library \texttt{stats}). The empirical discrete (\emph{e.g.} for
bootstrap), the Dirichlet, the multinomial and the multivariate normal
may be used with uncertain and/or variable parameters by specifying
multivariate nodes. See section \ref{sec:Multivar}.

@@

Another way to construct a \texttt{mcnode} object is \emph{via} the
\texttt{mcdata} function, when data are available.

@@

\texttt{mcnode}s can be automatically constructed using operations
on other \texttt{mcnode}s. Rules are used to transfer uncertainty
and variability coherently within the model. Logically, the rules
are as follows (illustrated here with a \texttt{+}, 
see table \ref{tab:Ops})\footnote{These rules are not the standard R rules for recycling.}:
\begin{itemize}
\item \texttt{0} + \texttt{0} = \texttt{0};
\item \texttt{0} + \texttt{V} = \texttt{V}
\item \texttt{0} + \texttt{U} = \texttt{U};
\item \texttt{0} + \texttt{VU}= \texttt{VU};
\item \texttt{V} + \texttt{V} = \texttt{V};
\item \texttt{V} + \texttt{U} = \texttt{VU} \footnote{the \texttt{U}
mcnode is recycled by row, the \texttt{V}
mcnode is recycled in the standard manner by column.};
\item \texttt{V} + \texttt{VU} = \texttt{VU} \footnote{the \texttt{V}
mcnode is recycled in the standard manner by column.};
\item \texttt{U} + \texttt{U} = \texttt{U};
\item \texttt{U} + \texttt{VU} = \texttt{VU} \footnote{the \texttt{U}
mcnode is recycled by row.};
\item \texttt{VU} + \texttt{VU} = \texttt{VU}
\end{itemize}

\begin{table}
\caption{\label{tab:Ops}\texttt{mcnode}s combinations}

\centering{}\begin{tabular}{|c||c|c|c|c|}
\hline 
 & 0 & V & U & VU\tabularnewline
\hline
\hline 
0 & 0 & V & U & VU\tabularnewline
\hline
V & V & V & VU & VU\tabularnewline
\hline 
U & U & VU & U & VU\tabularnewline
\hline 
VU & VU & VU & VU & VU\tabularnewline
\hline
\end{tabular}
\end{table}

Thus, in our example:

@@

\texttt{expo} is a function of a \texttt{U} and two \texttt{V}
\texttt{mcnode}s: it is a \texttt{VU} \texttt{mcnode} with
variability in the row dimension and uncertainty in the column dimension. 
\texttt{risk} is a function of a \texttt{U} and a \texttt{VU}
\texttt{mcnode}: it is therefore a \texttt{VU} \texttt{mcnode}.

@@

Special functions \texttt{is.na(x)}, \texttt{is.nan(x)}, \texttt{is.finite(x)},
\texttt{is.infinite(x)} are implemented to test if any values are
\texttt{NA} (missing data), \texttt{NaN} (\emph{Not A Number}),
finite or infinite .

@@

\footnote{Note that the resulting correlation (around 0.4) is obviously an approximation
to the desired value of 0.5, because a discrete distribution (\texttt{cook}:
3 categories) is correlated with a continuous distribution (\texttt{serving}).}

@@

One can re-run the model with different dimensions or random seeds in
one line:

@@
